---
layout: post
title: The death of Saturday football?
subtitle: The effects of declining fixture density on reducing the number of matchday goals in the Premier League
date: "`r Sys.Date()`"
tags: [R, soccer]
output: 
  md_document:
    variant: markdown_github
    preserve_yaml: true
bigimg: /img/motd.jpg
share-img: /img/motd.jpg
---

My [last post](https://jogall.github.io/2017-05-24-final-gameweek-extravaganzas/) looked at the highest number of goals scored in the final fixtures of a season, but this got me thinking what about the highest scoring single **matchday**, a.k.a. the best Match of the Day ever? In this post I'll examine showing a decreasing numbers of goals per matchday in the top division over recent decades and look at the effects of declining 'fixture density' (number of games played per day) to see whether fixtures are becoming more 'spread out' and if we can blame it on the death of Saturday fixture.

<!--- R configuration and post-specific webpage code --->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- add javascript to show hidden R code here -->
<!--- end --->

-----

To begin, we'll subset the top tier of the `england` dataframe from the R package `engsoccerdata` as in the last post:

```{r message=FALSE}
devtools::install_github("jalapic/engsoccerdata")
library(engsoccerdata)
library(dplyr)
library(ggplot2)

# Update 'england' dataframe if there are new results
england <- rbind(england, subset(england_current(), !(Date %in% england$Date & home %in% england$home)))

#subset all-time top flight and prettify the season variable for plotting (e.g. '2016' -> '2016-17')
topflight <- subset(england, tier==1) %>%
  arrange(Date) %>%
  mutate(season = as.factor(paste0(Season, "-", substr(Season+1, 3, 4))))
```

-----

###Best ever Match Of The Day?

First the easiest question: most goals scored in a single day?

```{r}
d1 <- topflight %>%
  group_by(Date) %>%
  summarise(Season = Season[1], goals = sum(totgoal), matches = n()) %>%
  arrange(-goals)
```

```{r echo=FALSE}
head(d1, 10)
```

The 66 goals gifted to us on Boxing Day 1963, and just look at those results:

```{r}
subset(topflight, Date=="1963-12-26") %>%
  select(Date, home, visitor, FT)
```

The only thing is, MotD wasn't on the air then - although it did start only 8 months later so it's possible this Boxing Day bonanza helped push the BBC towards televising football highlights. Before we continue, I still feel obliged to answer the original question and find out what is the highest scoring matchday since MotD began, i.e. post-August 1964?

```{r}
topflight %>%
  subset(Date >= "1964-08-22") %>%
  group_by(Date) %>%
  summarise(goals = sum(totgoal), matches = n()) %>%
  arrange(-goals) %>%
  head(10)
```

The slightly-less-exciting sum of 56 goals scored in December of the 1965-66 season:

```{r}
subset(topflight, Date=="1965-12-11") %>%
  select(Date, home, visitor, FT)
```

-----

###So teams are scoring less nowadays, right?

The first thing I notice when looking at the top table is that almost all of the highest scoring matchdays are from over 50 years ago; in fact, you have to look all the way down to the 83rd highest entry to find a fixture from the 1970s or later. If we plot the data we can visualise this decline in goals per matchday (red line shows a linear regression fitted to the data, blue line a [smoothed Loess curve](https://en.wikipedia.org/wiki/Local_regression)):

JAVASCRIPT_CODE_HERE
```{r dpi=300}
#set default theme and increase font size for all plots
theme_set(theme_bw(base_size = 14))

ggplot(d1, aes(x = Date, y = goals)) +
  geom_point(alpha=0.2) +
  geom_smooth(method='loess', se=FALSE, col="blue", alpha=0.5) +
  geom_smooth(method='lm', se=FALSE, col="red", alpha=0.5) +
  scale_y_continuous(lim=c(0,70)) +
  ylab("Total goals on matchday")
```
END_DIV_TAG

So why might this be happening? The most obvious explanation would be that there are simply fewer goals scored nowadays; after all, the modern game has evolved a lot in the last few decades in terms of both physicality and tactics. Let's check to see if the data supports this idea:

```{r}
d2 <- topflight %>%
  group_by(Season) %>%
  group_by(Season) %>%
  summarise(goals.sum = sum(totgoal), goals.per.game = sum(totgoal) / n(), games = n())
```

The total number of goals scored has definitely declined since the 1950s...

JAVASCRIPT_CODE_HERE
```{r dpi=300}
ggplot(d2, aes(x = Season, y = goals.sum)) +
  geom_point() +
  geom_smooth(method='loess', col="blue") +
  scale_x_continuous(breaks=c(1888, 1900, 1925, 1950, 1975, 2000, 2016)) +
  ylab("Total goals per season")
```
END_DIV_TAG

...but standardising for the amount of games played per season shows this decline is less dramatic in terms of goals per game.

JAVASCRIPT_CODE_HERE
```{r dpi=300}
ggplot(d2, aes(x = Season, y = goals.per.game)) +
  geom_point() +
  geom_smooth(method='loess', se=FALSE, col="blue", alpha=0.5) +
  geom_smooth(method='lm', se=FALSE, col="red", alpha=0.5) +
  scale_y_continuous(lim=c(0,5)) +
  scale_x_continuous(breaks=c(1888, 1900, 1925, 1950, 1975, 2000, 2016)) +
  ylab("Goals per game")
```
END_DIV_TAG

In fact, it looks like the period from the 1950s to the mid-1960s was exceptionally high scoring, but that things have been relatively stable since the 1970s (there's actually a trend for a slight increase):

JAVASCRIPT_CODE_HERE
```{r dpi=300}
ggplot(subset(d2, Season >= 1970), aes(x = Season, y = goals.per.game)) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE, col="red", alpha=0.5) +
  scale_y_continuous(lim=c(0,5)) +
  scale_x_continuous(breaks = c(1970, 1980, 1990, 2000, 2010, 2016)) +
  ylab("Goals per game")
```
END_DIV_TAG

So the number of goals per matchday looks to be decreasing but the number of goals per game doesn't? Surely then that suggests there are fewer **games per matchday** in the modern game compared to the pre-1970s... After all, haven't we seen the [death of the 3pm Saturday fixture in the Premier League?](https://qz.com/867361/why-premier-league-soccer-games-cant-be-shown-at-3pm-in-the-uk/) [^1] It seems to me nowadays that most weekends in the EPL feature only a handful of Saturday fixtures with a few being played instead on Sunday and sometimes Monday and/or Friday too. (The penultimate week of this season saw games played on Friday, Saturday, Sunday, Monday, Tuesday, Wednesday and Thursday!)

-----

###Fixture density

I can't recall ever seeing any data to support this idea though so let's investigate using our historical results. A straightforward way of measuring whether fixtures have become more 'spread out' could be to calculate a '**fixture density score**' by dividing the expected number of games per season by the number of unique matchdays that season. So in the modern 20-team EPL featuring 38 gameweeks, the highest score would be 10.0 if all 10 matches each gameweek were played on a Saturday ($380 / (38 \times 1) = 380 / 38 = 10.0$), and the lowest possible score would be 1.4 if the 10 matches each gameweek were spread out over all 7 days that week ($380 / (38 \times 7) = 380 / 266 = 1.43$). To make it more intuitive, let's divide this number by the expected number of matches per gameweek (in the example above, 10 games per week) to make our score between a minimum of 0 and maximum of 1 [^2].

```{r}
fixdens <- england %>% 
  group_by(division, Season) %>%
  summarise(gameweeks = (n_distinct(home)-1) * 2,
            matchdays = n_distinct(Date)) %>%
  mutate(fixdens = gameweeks / matchdays)
```

Let's visualise fixture density over the years in the top fight first:

JAVASCRIPT_CODE_HERE
```{r dpi=300}
ggplot(subset(fixdens, division == 1), aes(x = Season, y = fixdens)) +
  geom_point() + 
#  geom_smooth(method='lm', col="red") +
  geom_smooth(method='loess') +
  scale_x_continuous(breaks = c(1888, 1920, 1950, 1980, 2016)) +
  ylab("Fixture density")
```
END_DIV_TAG

There looks to be a trend of decreasing fixture density since the 1950s, and a particularly sharp dip since around 1992. Is this something to do with TV rights and increasing commercialisation of the top division in England? Let's have a look at fixture density in the lower divisions to compare [^3] [^4]:

```{r}
fixdens <- fixdens %>%
  mutate(division2 = recode(division, "3a" = "3", "3b" = "3"))
```

JAVASCRIPT_CODE_HERE
```{r dpi=300}
ggplot(fixdens, aes(x = Season, y = fixdens)) +
  geom_point() + 
  geom_smooth(method='lm', formula=y ~ poly(x, 3, raw=TRUE)) +
  # stat_smooth(method = "gam", formula = y ~ s(x, k = 3)) +
  scale_x_continuous(breaks = c(1888, 1920, 1950, 1980, 2016)) +
  ylab("Fixture density") +
  facet_wrap(~ division2)
```
END_DIV_TAG

It looks like the second tier (present-day EFL Championship) has experienced a similar decline in fixture density since the 1950s, although this seems to be offset by increased fixture density in the 3rd and 4th tiers.

What about the distribution of matchdays across **days of the week** - can we see whether this decline come from a decrease in Saturday fixtures? We can use the `wday()` function from the handy [`lubridate`](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html) package to infer days of the week from our dates.

```{r}
matchdays <- england %>%
  mutate(day = lubridate::wday(Date, label=TRUE)) %>%
  group_by(division, Season) %>%
  mutate(games = n()) %>%
  group_by(division, Season, day) %>%
  summarise(prop = n() / games[1], games = n()) %>%
  mutate(division2 = recode(division, "3a" = "3", "3b" = "3"))
```

JAVASCRIPT_CODE_HERE
```{r dpi=300}
ggplot(matchdays, aes(x = Season, y = prop, colour = day)) +
  geom_line() +
  scale_x_continuous(breaks = c(1888, 1920, 1950, 1980, 2016)) +
  scale_y_continuous(lim=c(0,1)) +
  ylab("Proportion of total games") +
  facet_wrap(~division2)
```
END_DIV_TAG

It looks like there is a decline in the proportion of matches played on a Saturday in the top two divisions; in the top flight, from around three-quarters in the 1950s to around half now. And this seems to be accounted for by an increased proportion of Sunday fixtures in the top division and Tuesday fixtures in the second division. The lower leagues seem to tell the same story except with an increasing proportion of Saturdays being replaced by Tuesdays, but I'm thinking that this is probably due to an increased occurrance of two fixtures per week due to the increased number of teams in their leagues. So let's look at the **absolute number** of games instead of proportions:

JAVASCRIPT_CODE_HERE
```{r dpi=300}
ggplot(matchdays, aes(x = Season, y = games, colour = day)) +
  geom_line() +
  scale_x_continuous(breaks = c(1888, 1920, 1950, 1980, 2016)) +
  ylab("Absolute number of games") +
  facet_wrap(~division2)
```
END_DIV_TAG

Just as I suspected: there appears to be a greater number of total number in the lower leagues but the number of Saturday fixtures in the top division has definitely been declining since around 1980 - although we can see now this is not happened in the second tier.

-----

###In conclusion...

So whilst the number of goals per game is much the same in the EPL now as it was during the First Division in the 1970s, the absolute number of goals - the important thing - scored per matchday is decreasing as a result of decreasing fixture density, most noticeable fewer Saturday fixtures. Here's the last figure to show that Saturday's Match Of The Day is offering up less goals.

```{r}
#subset Saturday fixtures only
d3 <- topflight %>%
  mutate(day = lubridate::wday(Date, label=TRUE)) %>%
  subset(day == "Sat") %>%
  group_by(Season) %>%
  summarise(games = n(), goals = sum(totgoal))
```

JAVASCRIPT_CODE_HERE
```{r dpi=300}
ggplot(d3, aes(x = Season, y = goals)) +
  geom_point() +
  geom_smooth(method='loess', col="blue") +
  scale_x_continuous(breaks=c(1888, 1900, 1925, 1950, 1975, 2000, 2016)) +
  ylab("Total Saturday goals")
```
END_DIV_TAG

-----

[^1]: Premier League fixtures kicking off between 3pm-5pm on a Saturday are now the only top flight fixtures not to be televised live in the UK, a blackout introduced by the Football League in the 1960s which aimed to protect attendances at lower league games.

[^2]: Here's the equation more explicitly: for leagues consisting of {x} number of teams, fixture density is the number of games per season ($2(x - 1) \cdot \frac{x}{2}$) divided by the number of unique matchdays ($m$) divided by the expected number of games per gameweek ($\frac{x}{2}$). This simplifies down to expected number of gameweeks per season divided by the number of unique matchdays per season ($\frac{2(x - 1)}{m}$).

[^3]: The roots of the third and fourth tiers of English football are [a bit of a nightmare](https://en.wikipedia.org/wiki/Football_League_Third_Division): the Third Division lasted only a single season before the league split into North and South divisions in 1921-22 and again merged in 1958-1959, with the top half of the league that season going on to form the new Third Division and the bottom half the new Fourth Division. To save on headaches, I've therefore combined division `3` (1920-21 Third Division + new Third Division) with `3a` (Third Division North) and `3b` (Third Division South), and left division 4 alone.

[^4]: I've sprung for a third order polynomial to fit the data as this curve appears to fit the data better and seems less biased by single outlyiers than a Loess curve.

[^5]: Spitting the season into gameweeks is somewhat troublesome due to some fixtures being rescheduled as the season wears on - for instance, one team's 32nd game might take place in the same week as another team's 33rd game. So I've taken some liberty by simply splitting the season into an equal number of chunks (e.g. 38, 10-match chunks for the 2016-17 season). This simplified method might mean that some 10 fixture chunks contain multiple games by the same team whilst excluding other teams.