---
layout: post
title: Boring, boring Man United? Can we measure excitement and does it correlate with success?
subtitle: Goals, net goalscoring opportunities, efficiency and luck
date: "2017-05-19"
tags: [R, soccer]
output: 
  md_document:
    variant: markdown_github
    preserve_yaml: true
bigimg: /img/yawn.jpg
---

The motivation for this post came from watching yet another of Man United's goalless games, this time against Southampton on Wednesday night. It's one part a serious examination of United's current trend of low scoring -- how this fits in with their previous Premier League record and whether 'negative football' explains their recent lack of success in the league -- and one part the schadenfreude of a Liverpool fan looking to see if I can pronounce them The Most Boring Team Ever.

_____
  
Let's fire up `R` and use the most recent build of the 'engsoccerdata' package to get all previous results in Premier League (1992-93 to present), as well as some packages for manipulating and visualisating the data. I won't include the code used to create the plots in this post but it can be found [here](https://github.com/JoGall/jogall.github.io/Rmd/2017-05-19-does-excitement-bring-success.Rmd).
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_chunk$set(fig.path = "/home/joe/Dropbox/github/jogall.github.io/assets/POST-TITLE_files/")
```

```{r message=FALSE}
devtools::install_github("jalapic/engsoccerdata")
require(engsoccerdata)
require(ggplot2)
require(dplyr)

#Update 'england' dataframe if needed
england <- rbind(england, subset(england_current(), !(Date %in% england$Date & home %in% england$home)))

#Subset to PL era and prettify the season variable for plotting (e.g. '2016' -> '2016-17')
EPL <- subset(england, tier == 1 & Season %in% 1992:2016) %>%
  mutate(season = as.factor(paste0(Season, "-", substr(Season+1, 3, 4))))
```

First up, let's look at United's goalscoring record this season using the `maketable_all()` function:
```{r}
#League goals this season
maketable_all(subset(EPL, Season == 2016 & tier == 1)) %>%
  select(team, P = GP, F = gf, A = ga, GD = gd, "(Pos)" = Pos) %>%
  arrange(-F)
```

They rank 8th with 54 goals scored -- noticeably the lowest of the 'top 6'; even Bournemouth have scored more than them this season.

### How to quantify 'boring'?

Thinking along the lines of measuring 'boring', what about making some kind of 'excitement score'? For neutral viewers, I think the total number of goals scored in a game is often a decent measure for how exciting it was to watch.

```{r}
maketable_all(subset(EPL, Season == 2016 & tier == 1)) %>%
  mutate(goals = gf + ga, goals_per_game = round( (gf + ga)/ GP, 2)) %>%
  select(team, goals, goals_per_game) %>%
  arrange(-goals)
```

United rank 19th in this table, their 38 games offering up 83 goals. Only Middlesbrough (80) have provided less excitement for the netural, and even West Brom have produced more than United (no offence, Baggies). Bournemouth top this league with nearly 50% more excitement than United! (Their goal difference (-12) suggests things are probably a bit less exciting if you're a Bournemouth fan though.)

Also, we can't forget the thing that everyone seems to be saying has epitomised United's season: the bore-draw; the nil-nil; the neutral's dreaded 0-0. (Although I know there have been plenty of examples of scintillating nil-nils; see [footnote here](#footnote).)

```{r}
subset(EPL, Season == 2016 & tier == 1) %>%
  homeaway() %>%
  subset(gf==0 & ga==0) %>%
  group_by(team) %>%
  summarise(nil_nils = n()) %>%
  arrange(-nil_nils) %>%
  head(5)
```

United have given us 6 nil-nils this season; only Boro and Southampton have offered more with by going one further.

It's not all bad though, more clean sheets are an obvious perk to this trend of shut-outs and nil-nils -- United sit joint-top of the clean sheet league with Spurs this season (17):

```{r}
subset(EPL, Season == 2016 & tier == 1) %>%
  homeaway() %>%
  subset(ga==0) %>%
  group_by(team) %>%
  summarise(clean_sheets = n()) %>%
  arrange(-clean_sheets) %>%
  head(5)
```


_____


### Is this a new trend for United?

But how does this compare with previous seasons for Man Utd?

```{r}
d1 <- lapply(unique(EPL$season), function(x) {
  maketable_all(subset(EPL, season == x & tier == 1)) %>%
    mutate(season = x)
}) %>%
plyr::rbind.fill() %>%
subset(team == "Manchester United")

d2 <- lapply(1992:2016, function(x) {
  homeaway(subset(EPL, Season == x & tier == 1)) %>%
    subset(team == "Manchester United")
}) %>%
  plyr::rbind.fill() %>%
  group_by(Season) %>%
  summarise(g_p90 = sum(gf+ga) / n(), 
            gf_p90 = sum(gf) / n(), 
            ga_p90 = sum(ga) / n(), 
            no_score = sum(gf==0), 
            clean_sheets = sum(ga==0), 
            nil_nils = sum(gf==0 & ga==0)
            ) %>%
  mutate(season = as.factor(paste0(Season, "-", substr(Season+1, 3, 4))))
```

Let's look at our throwaway excitement score first. So this season was United's 'least exciting' in terms of goals per game, although only marginally lower than previous record shared between the 2015-16 season and the 2004-05 season.

```{r, echo=FALSE, dpi=300}
ggplot(d2, aes(x = season, y = g_p90, group=1)) + 
  geom_point(stat='summary', fun.y=sum, col="red") +
  stat_summary(fun.y=sum, geom="line", col="red") +
  ylab("Total goals per game") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

2004-05 was also the only other season that produced as many nil-nils.

```{r, echo=FALSE, dpi=300}
ggplot(d2, aes(x = season, y = nil_nils, group=1)) + 
  geom_point(stat='summary', fun.y=sum, col="red") +
  stat_summary(fun.y=sum, geom="line", col="red") +
  ylab("Nil-nils") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

Though it has to be said that United don't even approach the record for most boring team in a Premier League season though.

```{r}
d3 <- lapply(1992:2016, function(x) {
  homeaway(subset(EPL, Season == x & tier == 1))
}) %>%
  plyr::rbind.fill() %>%
  group_by(Season, team) %>%
  summarise(g_p90 = sum(gf+ga) / n(), 
            gf_p90 = sum(gf) / n(), 
            ga_p90 = sum(ga) / n(), 
            no_score = sum(gf==0), 
            clean_sheets = sum(ga==0), 
            nil_nils = sum(gf==0 & ga==0)
            ) %>%
  arrange(g_p90)

which(d3$team=="Manchester United" & d3$Season == 2016)
```

Their 2016-17 campaign ranks 32nd in the lowest number of total number of goals per game, and there have been plenty of teams producing more goalless games in a season. Leeds United can proudly hold both records for lowest excitement delivered: their 1996-97 season produced a snooze-worthy 1.74 goals per game on average and **9** nil-nils.

A bit more seriously, let's focus on United's own goalscoring. The last two seasons have saw them score the fewest goals per game in their EPL history...

```{r, echo=FALSE, dpi=300}
ggplot(d2, aes(x = season, y = gf_p90, group=1)) + 
  geom_point(stat='summary', fun.y=sum, col="red") +
  stat_summary(fun.y=sum, geom="line", col="red") +
  ylab("Total goals per game") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

...and the apparent silver lining provided by keeping more clean sheets doesn't seem so great compared to previous seasons.

```{r, echo=FALSE, dpi=300}
ggplot(d2, aes(x = season, y = clean_sheets, group=1)) + 
  geom_point(stat='summary', fun.y=sum, col="red") +
  stat_summary(fun.y=sum, geom="line", col="red") +
  scale_y_continuous(lim=c(10, 25)) +
  ylab("Clean sheets") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

Simply put, United have scored more whilst conceding less in previous campaigns. Here's a plot of both goals for and against over the years, just for the craic.

```{r, echo=FALSE, dpi=300}
ggplot(d1, aes(x = season, y = gf / GP, group=1)) + 
  geom_point(aes(x = season, y = gf / GP, group=1, colour="For"), stat='summary', fun.y=sum) +
  stat_summary(fun.y=sum, geom="line", col="red") +
  geom_point(aes(x = season, y = ga / GP, group=1, colour="Against"), stat='summary', fun.y=sum) +
  stat_summary(aes(x = season, y = ga / GP, group=1), fun.y=sum, geom="line", col="blue") +
  scale_y_continuous(lim=c(0,3), expand=c(0,0)) +
  ylab("Mean goals per game") +
  xlab("") +
  scale_colour_manual(name="", values = c("For" = "red", "Against" = "blue")) +
  scale_linetype_manual(name="Legend", values = c("a" = "dashed", "b" = "dotted")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1),
        legend.position=c(0.95,0.05), legend.justification=c(1,0))
```

_____

### Going deeper than goals: shots for and against

I want to wade a bit deeper into match statistics for a bit -- specifically shots for / against and shots on target for / against -- to try and see whether we can show that less attacking play is what is behind this apparent goal drought.

Nothing near as fancy as the stats we might find on www.squawka.com or similar then, but we'll need some more detailed data as this information isn't contained in 'engsoccerdata' package. We can use the website [Football-Data](http://www.football-data.co.uk/data.php), which contains historical data on matches from several European leagues. It takes a bit of effort to clean up but the home-made function 'getFD()' gets us there.

```{r}
#make season codes for URLs
getFD <- function(start = NULL, end = NULL) {
  
  #make URLs
  URLs <- paste0("http://www.football-data.co.uk/mmz4281/",
                 paste0(substr(start:end, 3, 4), substr((start+1):(end+1), 3, 4)),
                 "/E0.csv")
  
  lapply(URLs, function(x) {
    
    #get CSV for each season
    mydf <- read.csv(x)
    
    #clean up empty rows
    mydf <- mydf[mydf$Date!="",]
    
    #conform YYYY dates to YY (ifelse makes Date variables numeric, so need to wrap in as.Date and specify origin)
    mydf$Date <- as.Date(
                    ifelse(nchar(as.character(mydf$Date))==10, 
                                as.Date(mydf$Date, format="%d/%m/%Y"),
                                as.Date(mydf$Date, format="%d/%m/%y")
                     ), origin = as.Date("2015-01-01") - as.numeric(as.Date("2015-01-01")))
    
    #add 'Season' variable as year of first fixture
    mydf$Season <- as.numeric(format(mydf$Date[1], "%Y"))
    
    mydf
  }) %>%
    plyr::rbind.fill()
}
```

We'll use this function to get the shots statistics we're after: shots for ('sf'), shots against ('sa'), shots on target for ('sotf') and shots on target against ('sota'). We'll normalise these to games played to account for the fact there's still one game to play this season. (Note I've only gone as far back as the 2000-01 season as that's earliest season to contain this data.) We'll reformat it in the 'homeaway()' style used by 'engsoccerdata', which let's us look at all results from each team's perspective.

```{r}
#get raw data
shots_raw <- getFD(2000, 2016)

#convert to 'homeaway()' format
shots <- rbind(
  shots_raw %>%
    select(Season, Date, team = HomeTeam, opp = AwayTeam, gf = FTHG, ga = FTAG, result = FTR, sf = HS, sa = AS, sotf= HST, sota = AST) %>% 
    mutate(venue = "home"), 
  shots_raw %>%
    mutate(away_result = ifelse(FTR == "H", "A", ifelse(FTR == "A", "H", "D"))) %>%
    select(Season, Date, team = AwayTeam, opp = HomeTeam, gf = FTAG, ga = FTHG, result = away_result, sf = AS, sa = HS, sotf= AST, sota = HST) %>% 
    mutate(venue = "away")) %>%
  arrange(team, Date) %>%
  mutate(season = as.factor(paste0(Season, "-", substr(Season+1, 3, 4))) )

head(shots, 5)
```

So now we've got the data cleaned up an ready to go, let's look at United's previous records for shots for and against. First, shots on goal was at its lowest last season but looks back to normal this season.

```{r, echo=FALSE, dpi=300}
ggplot(subset(shots, team=="Man United"), aes(x = season, y = sf)) + 
  stat_summary(fun.y = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  ylab("Shots per game") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

Shots on target tells a different story, showing a real difference between United over the last 4 seasons and their more dominant period between 2000-01 to 2012-13.

```{r, echo=FALSE, dpi=300}
ggplot(subset(shots, team=="Man United"), aes(x = season, y = sotf)) + 
  stat_summary(fun.y = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  ylab("Shots on target per game") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

Again though, this seems to work both ways - United have also afforded their opposition fewer goalscoring opportunities in the last 4 seasons.

```{r, echo=FALSE, dpi=300}
ggplot(subset(shots, team=="Man United"), aes(x = season, y = sota)) + 
  stat_summary(fun.y = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  ylab("Shots on target against per game") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

Although expanding our 'excitement score' from total goals per game to total shots per game, these 4 seasons look **much** less exciting for the neutral.

```{r, echo=FALSE, dpi=300}
ggplot(subset(shots, team=="Man United"), aes(x = season, y = sotf + sota)) + 
  stat_summary(fun.y = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  ylab("Shots on target per game") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

_____

### 'Net goalscoring opportunities'

I want to do something more interesting with our shots data. I'm interested in whether this so-called 'negative football' -- creating less but also stifling the opposition -- works or not. Let's make a metric called 'net goalscoring opportunities': simply the difference between chances created and chances conceded. Here's the data for United's net shots per game over previous seasons (I won't show the data for shots on target but it looks very similar).

```{r, echo=FALSE, dpi=300}
ggplot(subset(shots, team=="Man United"), aes(x = season, y = sf - sa)) + 
  stat_summary(fun.y = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  ylab("Net goalscoring opportunites (shots for - shots against)") +
  xlab("") +
  geom_hline(yintercept=0, lty=3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

Last season was by far the worst, entering the negative in some games, but this season's United look back to old form. I'm not sure how useful this number is in measuring success; United's 2nd lowest score came in the 2012-13 season yet they won the league..

Let's investigate the value of the metric further by trying to find a team we'd expect to have a net goalscoring opportunity. A team that's got a few PL seasons under their belt but spent much time in the bottom half of the league maybe. West Brom are the first team that come to mind (sorry again, Baggies).

```{r, echo=FALSE, dpi=300}
ggplot(subset(shots, team=="West Brom"), aes(x = season, y = sf - sa)) + 
  stat_summary(fun.y = mean, geom = "point") +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  ylab("Net goalscoring opportunites (shots for - shots against)") +
  xlab("") +
  geom_hline(yintercept=0, lty=3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

This follows what we expected. How about seeing whether net goalscoring opportunities (I'm refusing to acronym-ise this) is a decent predictor of results by plotting it against final league position. We need to dip back into 'engsoccerdata' to make league tables for each season, and use a touch of wizardry to merge these dataframes together as they contain different team name variations (e.g. 'Man City' vs. 'Manchester City')

```{r}
#summarise shots data
ss <- shots %>%
  group_by(Season, team) %>%
  mutate(GP = n() ) %>%
  summarise(net_S = (sum(sf) / n()) - (sum(sa) / n()), net_SoT  = (sum(sotf) / n()) - (sum(sota) / n()) )

#make sure team names in our new dataset match those used in engsoccerdata
ss$team <- teamnames$name[match(ss$team, teamnames$name_other)]

#calculate league tables
tables <- lapply(unique(ss$Season), function(x) {
  maketable(england, Season = x, tier = 1) %>%
    mutate(Season = x, Pos = as.numeric(Pos)) %>%
    dplyr::select(Season, Pos, team, gf, ga, gd)
}) %>%
plyr::rbind.fill()

#merge dataframes
ss2 <- merge(ss, tables, c("Season", "team")) %>%
  mutate(season = as.factor(paste0(Season, "-", substr(Season+1, 3, 4))) )
```

Let's plot it: net total shots vs. league position...

```{r, echo=FALSE, dpi=300}
ggplot(ss2, aes(x = Pos, y = net_S)) + 
  geom_point(alpha=0.5) +
  geom_smooth(method=lm) +
  geom_point(data = subset(ss2, team=="Manchester United" & Season == 2016), aes(x = Pos, y = net_SoT), col="red", size=4) +
  geom_hline(yintercept=0, lty=3) +
  scale_x_reverse(breaks=c(20, 15, 10, 5, 1), limits=c(20, 1)) +
  ylab("Net shots (for - against)") +
  xlab("League position") +
  theme_bw()
```

...and net shots on target vs. league position:

```{r, echo=FALSE, dpi=300}
ggplot(ss2, aes(x = Pos, y = net_SoT)) + 
  geom_point(alpha=0.5) +
  geom_smooth(method=lm) +
  geom_point(data = subset(ss2, team=="Manchester United" & Season == 2016), aes(x = Pos, y = net_SoT), col="red", size=4) +
  geom_hline(yintercept=0, lty=3) +
  scale_x_reverse(breaks=c(20, 15, 10, 5, 1), limits=c(20, 1)) +
  ylab("Net shots on target (for - against)") +
  xlab("League position") +
  theme_bw()
```

It looks like a decent metric of success. In both cases, the line of best fit crosses the zero line at around 10-11th position, exactly where we'd have expected when trying to separate the top and bottom teams.

Not forgetting Man United, I've added their point for this season in red. They're placed exactly as expected considering their net shots but actually have the best net shots on target of any team finishing in 6th (+2.6 SoT per game) and could even be considered unlucky: team's have won the league with similar figures in other seasons.

But hang on; this pattern is almost identical to what we see with goal difference...

```{r, echo=FALSE, dpi=300}
ggplot(tables, aes(x = Pos, y = gd)) + 
  geom_point(alpha=0.5) +
  geom_smooth(method=lm) +
  geom_point(data = subset(ss2, team=="Manchester United" & Season == 2016), aes(x = Pos, y = net_SoT), col="red", size=4) +
  scale_x_reverse(breaks=c(20, 15, 10, 5, 1), limits=c(20, 1)) +
  geom_hline(yintercept=0, lty=3) +
  ylab("Goal difference") +
  xlab("League position") +
  theme_bw()
```

...and this similarity might be due to a strong correlation between net goalscoring opportunites and actual goal difference. Not that that should be too surprising.

```{r, echo=FALSE, dpi=300}
ggplot(ss2, aes(x = net_SoT, y = gd)) + 
  geom_point(alpha=0.5) +
  geom_smooth(method=lm) +
  geom_point(data = subset(ss2, team=="Manchester United" & Season == 2016), aes(x = net_SoT, y = gd), col="red", size=4) +
  geom_text(aes(x = 3.5, y = 15), label = "Man Utd (2016-17)") +
  geom_hline(yintercept=0, lty=3) +
  geom_vline(xintercept=0, lty=3) +
  ylab("Actual goal difference") +
  xlab("Net goalscoring opportunities (SoT for - SoT against)") +
  theme_bw()
```

_____

### Efficiency or luck?

I'm not convinced our 'net goalscoring opportunites metric offers any advantage over straight-up goal difference given how correlated the two are, although it's possible the distance of an individual point from the line of best fit could tell us something worthwhile. Teams above the line have a better goal difference than expected considering the balance between the number of goalscoring opportunities they create and the concede. This seems to be sort of measure of 'effectiveness' (I'm not sure what to better call it): how good teams are at converting chances into goals, and how well they deal with opponents' chances. Teams falling above the line might be great at burying their chances and/or have a great shot-stopper in goal.

It might also be aruged that the deviations from this line of best fit are often considered good luck or bad luck. Teams below the line convert a lower proportion of their shots into goals than expected and/or concede more goals than expected considering the number of shots they face.

Let's just naively call it 'luck' for now seeming as we don't know much yet. We'll measure it as the residual of a point, i.e. its distance from a linear regression fitted to the above data; negative scores are 'unlucky' teams which fall below the line and vice versa.

```{r}
#fit a linear regression
ss3 <- na.omit(ss2)
mod1 <- lm(gd ~ net_SoT, ss3)
#measure luck as each point's residual
ss3 <- ss3 %>%
  mutate(luck = residuals(mod1)) %>%
  arrange(luck)
```

Here's the highest 5 / lowest 5:

```{r, echo=FALSE}
rbind(head(ss3, 5), tail(ss3, 5))
```

Interestingly, the first team that came to mind was Leicester this season prior to Ranieri's departure. I remember hearing a stat that they'd been unlucky in the season so far (although I can't recall how this was quantified) and were due a good run, and the next week they went and beat Liverpool 3-1 (typical). But taking their position from the results before game (yellow point, below), they actually rank above the line.

The unluckiest and/or most wasteful and/or most leaky defensive season in PL history was had by Wigan in 2009-10, who fall way below the line (dark blue). And the flukiest and/or most clinical and/or defensive rock of a season was had by Man City in 2013-14 (sky blue). Not sure this stat makes me feel any better about that season whatever it means, but I'm sure it's a nice bonus for any United fans that are still reading.

```{r, echo=FALSE, dpi=300}
leicester_point <- shots %>%
  subset(team == "Leicester" & Date > "2016-08-13" & Date <= "2017-02-27") %>%
  summarise(gd = sum(gf) - sum(ga), net_S = (sum(sf) / n()) - (sum(sa) / n()), net_SoT  = (sum(sotf) / n()) - (sum(sota) / n()) )

ggplot(ss2, aes(x = net_SoT, y = gd)) + 
  geom_point(alpha=0.5) +
  geom_smooth(method=lm) +
  geom_point(data = subset(ss2, team=="Manchester City" & Season == 2013), aes(x = net_SoT, y = gd), col="lightblue", size=4) +
  geom_point(data = subset(ss2, team=="Wigan Athletic" & Season == 2009), aes(x = net_SoT, y = gd), col="darkblue", size=4) +
  geom_point(aes(x = leicester_point$net_S, y = leicester_point$net_SoT), col="yellow", size=4) +
  geom_hline(yintercept=0, lty=3) +
  geom_vline(xintercept=0, lty=3) +
  ylab("Actual goal difference") +
  xlab("Net goalscoring opportunities (SoT for - SoT against)") +
  theme_bw()
```

_____
  
### In conclusion...

I'll wrap this post up now. It's been another loosely-directed paddle in the shallows of the bottomless sea that is football statistics, but hopefully another example of what can be readily done with freely-available datasets, open-source software, and the art of creating probably-useless metrics that may or may not mean anything.

With regards to United, the stats do support the concesus that they've had a boring season and more generally appear to have been in a rather boring transitional period over the last 4 seasons. Obviously this isn't a new thing under Mourinho then, but as a Red that doesn't think he'll ever do the business at United, I say long may he reign. Or maybe I just love to hate the man.

Coming soon is a thrillogy of posts on 'Robin Hood' teams: those that take points from the top teams and redistribute them to the bottom, with an obvious focus on Liverpool this season, who are unbeaten against the top 6 yet have been woeful against the bottom half teams. I might take a more in-depth look at luck in a future post too; I'm particularly interested in finding a metric that can be used as a predictive tool, i.e. what team's recent form may just be down to bad luck and be worth a punt in the bookies?

_____


### Footnote:
What was the most exciting nil-nil ever? i.e. the game with the highest number of shots / shots on target that somehow ended goalless?

Funnily enough, Man United gave us The Most Exciting Nil-Nil Of All-Time (well, of the period 2000-01 - present) against Burnley at Old Trafford this season. That is if we go by total number of shots -- there were 44 shots in this game, 37 of those taken by United!

```{r}
subset(shots_raw, FTHG==0 & FTAG==0) %>%
  mutate(shots = HS + AS) %>%
  select(Date, HomeTeam, AwayTeam, shots, home = HS, away = AS) %>%
  arrange(-shots) %>%
  head(5)
```

Or if clinical excitement is more your thing, the title of Most-Shots-On-Target-In-A-Nil-Nil goes to the 25 SoT recorded in the game between Spurs and Man City at White Hart Lane two seasons ago; Spurs: 18, City: 7.

```{r}
subset(shots_raw, FTHG==0 & FTAG==0) %>%
  mutate(SoT = HST + AST) %>%
  select(Date, HomeTeam, AwayTeam, SoT, home = HST, away = AST) %>%
  arrange(-SoT) %>%
  head(5)
```

_____